---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(here)
here::i_am("README.md")

data_task_1 <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2020/2020-09-22/expeditions.csv")
data_task_2 <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2019/2019-02-19/phd_by_field.csv")
data_task_3 <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2019/2019-11-05/commute.csv")
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
data_task_1 |> 
  mutate(peak_name = forcats::fct_lump(peak_name, n = 15)) |>
  filter(peak_name != "Other") |>
  mutate(peak_name = forcats::fct_reorder(peak_name, peak_name, .fun = length)) |> 
  group_by(season, peak_name) |>
  ggplot(aes(x = peak_name, fill = season)) +
  geom_bar(aes(fill = season), show.legend = T) +
  coord_flip() +
  scale_fill_viridis_d() +
  theme_light() +
  labs(title = "The 15 most popular peaks stacked by season of expedition",
       y = "Number of expeditions") +
  theme(legend.position = "bottom",
        axis.title.y = element_blank(),
        plot.title = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.text = element_text(size = 12)
        )

ggsave(here("assignments", "assignment_2_plots", "task_1_expeditions.png"), width = 8, height = 6, dpi = 200)

```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
data_task_2 |> 
  group_by(broad_field, year) |> 
  summarise(total_phds = sum(n_phds, na.rm = TRUE)) |> 
  ggplot(aes(x = year, y = total_phds, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(labels = scales::comma_format()) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  labs(title = "Number of awarded Ph.D.-s in the US by year",
       color = "Broad field") +
  theme(legend.position = "right",
        plot.title = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        axis.text = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.title.x = element_blank()
        )

ggsave(here("assignments", "assignment_2_plots", "task_2_phds_awarded.png"), width = 8, height = 6, dpi = 200)
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
data_task_3 |> 
  group_by(state) |> 
  summarise(total_bike = sum(bike, na.rm = TRUE),
            total_foot = sum(foot, na.rm = TRUE)) |> 
  ggplot(aes(x = total_bike, y = total_foot)) +
  geom_point(size = 2, color = "steelblue") +
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format()) +
  theme_light() +
  labs(title = "Commuting to work by bike vs on foot in the US by state",
       x = "Total bike commutes",
       y = "Total foot commutes") +
  theme(plot.title = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12)
        )

ggsave(here("assignments", "assignment_2_plots", "task_3_commute.png"), width = 8, height = 6, dpi = 200)
```
